---
title: "01 - Simulating data"
date: "2019-04-30"
output: github_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include=FALSE} 
library(knitr)
```

Here is some code to set up simulated data sets. The main aim of these
simulations are to test the performance of phyloflows MCMC algorithm. Please
read the sections "Our Job" and "Our Solution" on the
main page before you go ahead here.

## Data set 1 (simple, SARWS)
We start with simulating transmission flows between two population groups called
"1" and "2". We will assume a very simple sampling process, sampling at random
within population strata, which we abbreviate to SARWS.

**We first set up the sampling process within the two population groups.** Let
us assume population group 1 consists of $2000$ individuals and group 2 of
$2500$ individuals, and that the sampling rates are $0.6$ for group 1 and
$0.45$ for group 2. Here, we will suppose that sampling is at random within
each of the population groups with these two sampling probabilities:  
```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE} 
library(data.table)
set.seed(42)
ds <- data.table(CATEGORY=c(1,2),TRIAL=c(2000,2500),P_SEQ_EMP=c(0.6,0.45))
ds[,SUC:=TRIAL*P_SEQ_EMP]
ds
```

**Next, we calculate the sampling probabilities of transmission flows
within and between the two groups.**  
```{r include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE} 
dobs <- data.table(TR_TRM_CATEGORY=c(1,1,2,2),REC_TRM_CATEGORY=c(1,2,1,2))
tmp <- subset(ds,select=c('CATEGORY','P_SEQ_EMP'))
setnames(tmp,colnames(tmp),paste0('TR_TRM_',colnames(tmp)))
dobs <- merge(dobs,tmp,by='TR_TRM_CATEGORY')
setnames(tmp,colnames(tmp),gsub('TR_','REC_',colnames(tmp)))
dobs <- merge(dobs,tmp,by='REC_TRM_CATEGORY')
dobs[, S:= TR_TRM_P_SEQ_EMP * REC_TRM_P_SEQ_EMP]
set(dobs, NULL, c('TR_TRM_P_SEQ_EMP','REC_TRM_P_SEQ_EMP'), NULL)
setkey(dobs,TR_TRM_CATEGORY,REC_TRM_CATEGORY)
dobs
```

**Next, we simulate true transmission flows.** Let us assume 36\% and 54\%
transmissions are within group 1 and 2 respectively, and 4\% are from group 1 to
group 2 and 6\% are from group 2 to group 1: 
$$
\pi=(0.36,0.04,0.06,0.54).
$$ 
We further assume the total number of observed transmissions is $N=300$. We will simulate the actual transmission count $Z$ from a Poisson
distribution. Then we will generate transmission flows between groups by 
$$
z \sim \mbox{Multinomial} (Z,\pi).
$$
Finally we will generate observed transmissions flows by subsampling the
actual transmission flows by
$$
n_{ab} \sim \mbox{Binomial} (z_{ab},\xi_{ab}), \forall a,b,
$$ 
where $\xi_{ab}$ is the probability of sampling a transmission event from $a$ to $b$.

```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE} 
TRUE_PI <- c(0.36,0.04,0.06,0.54)
N <- rpois(1,300/mean(dobs$S))
z <- rmultinom(1,size=N,prob=TRUE_PI)
n <- matrix(NA_integer_,ncol=1,nrow=length(TRUE_PI))
for (i in 1:length(TRUE_PI)){
	n[i] <- rbinom(1,size=z[i],dobs$S[i])
}
dobs[, TRM_OBS:= n]
dobs
```

## Corresponding **phyloflows** input data (simple, SARWS)
**We will start by bringing the transmission count data into the form needed for
phyloflows MCMC algorithm.** We add an ID to each observation, called
`TRM_CAT_PAIR_ID`. We also define the sampling groups. In this example, they
correspond directly to the transmission groups.
```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE} 
dobs[, TR_SAMPLING_CATEGORY:=TR_TRM_CATEGORY]
dobs[, REC_SAMPLING_CATEGORY:=REC_TRM_CATEGORY]
setkey(dobs,TR_TRM_CATEGORY,REC_TRM_CATEGORY)
dobs[, TRM_CAT_PAIR_ID:= seq_len(nrow(dobs))]
dobs[, S:=NULL]
dobs
```

**We also need to define the prior distribution on the unknown sampling
probabilities, and generate samples from it**. At the very top of this page,
defined the number of infected and sampled individuals in data.frame `ds`. Let
us denote these by $X_a^i$ and $X_a^s$ for our two population groups $a$. 
Usually this type of information is available to us in real-world
data analyses, and so we work from these numbers here also. Under the Binomial
sampling model that we assume throughout, $X_a^s\sim Binom(X_a^i, \xi_a)$. If we 
suppose a flat prior on $\xi_a$, we obtain the posterior distribution of the sampling 
probabilities conditional on the number of total and sampled individuals,
$$
p(\xi_a|X_a^i,X_s^i)= Beta(\xi_a;X_a^s+1,X_a^i-X_a^s+1).
$$
This density typically contains a lot of information on the sampling process.
To get this information into the form needed for **phyloflows** MCMC
algorithm, we need to derive samples from that distribution. We also need to
calculate their log density.
```{r include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE} 
nprior<-1000
dprior<-ds[,list(P=rbeta(nprior,SUC+1,TRIAL-SUC+1),
         SAMPLE=1:nprior),
   by='CATEGORY']
dprior<-merge(dprior,ds,by='CATEGORY')
dprior[,LP:=dbeta(P,SUC+1,TRIAL-SUC+1,log=TRUE)]
set(dprior,NULL,c('TRIAL','SUC'),NULL)
setnames(dprior, 'CATEGORY', 'SAMPLING_CATEGORY')
head(dprior)
```

This data set can be loaded through
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
data(twoGroupFlows1)
```
To test our MCMC algorithm, we repeated generated 100 such data set, and they
can be loaded through 
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
data(twoGroupFlows100)
```

We implemented our MCMC algorithm on these 100 data set
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
for (i in 1:100){
  dobs  <-  twoGroupFlows100[[i]][[1]]
  dprior_sarws  <-  twoGroupFlows100[[i]][[2]]
  dprior_sar  <-  twoGroupFlows100[[i]][[3]]
  randomnumber  <-  twoGroupFlows100[[i]][[4]]
  
  # number of iterations
  tmp  <- subset(dobs, select=c(TRM_CAT_PAIR_ID, TR_SAMPLING_CATEGORY, REC_SAMPLING_CATEGORY))
  tmp  <- melt(tmp, id.vars='TRM_CAT_PAIR_ID', value.name='SAMPLING_CATEGORY', variable.name='WHO')
  dlu            <- unique(subset(tmp, select=c(WHO, SAMPLING_CATEGORY)))
  dlu[, UPDATE_ID:= seq_len(nrow(dlu))]
  
  # mcmc
  mcmc.file  <-  paste0("sarws_mcmc",i,".rda")
  control  <-  list(seed=randomnumber, mcmc.n=nrow(dlu)*1e5, verbose=0, outfile=mcmc.file)
  source.attribution.mcmc(dobs, dprior_sarws, control)
  
  mcmc.file<-paste0("sar_mcmc",i,".rda")
  control<-list(seed=randomnumber, mcmc.n=nrow(dlu)*1e5, verbose=0, outfile=mcmc.file)
  source.attribution.mcmc(dobs, dprior_sar, control)
}
```

The estimates of $\pi$ is given by the medians from MCMC outputs. We calculated the mean absolute errors and worst case errors of estimation in order to validate the MCMC algorithm and compare the performance of SARWS and SAR approaches.

```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
# true PI
TRUE_PI  <-  c(0.36,0.04,0.06,0.54)

# record median of PI for each mcmc outputs
PI_M_SARWS_M <-  matrix(NA_real_,nrow = 100, ncol=4)
PI_M_SAR_M  <-  matrix(NA_real_,nrow = 100, ncol=4)

for (i in 1:100){
  load(paste0("sarws_mcmc",i,".rda"))
  # remove burnin
  burnin  <-  round(nrow(mc$pars$S)*0.2)
  tmp.pi  <-  1:nrow(mc$pars$PI)
  id.pi  <-  tmp.pi[tmp.pi>burnin]
  PI  <-  mc$pars$PI[id.pi,]
  # estimated PI
  PI_M_SARWS  <-  apply(PI, 2, median)
  PI_M_SARWS_M[i,]  <-  PI_M_SARWS
}

# calculate the mean absolute error and worst case error of estimation
PI_TRUE_M  <-  t(replicate(100, TRUE_PI))
PI_ABS_ERROR_SARWS  <-  abs(PI_M_SARWS_M-PI_TRUE_M)
PI_MAE_SARWS  <-  apply(PI_ABS_ERROR_SARWS,1,mean)
PI_WCE_SARWS  <-  apply(PI_ABS_ERROR_SARWS,1,max)
PI_MAE_SARWS.df  <-  data.table(PI_MAE=PI_MAE_SARWS, REPLICATE=1:100,
                                SCENARIO=rep(1,100), N=rep(300,100), SAMP_DIFF=rep(0.15,100))
PI_WCE_SARWS.df  <-  data.table(PI_WCE=PI_WCE_SARWS, REPLICATE=1:100, 
                                SCENARIO=rep(1,100), N=rep(300,100), SAMP_DIFF=rep(0.15,100))

for (i in 1:100){
  load(paste0("sar_mcmc",i,".rda"))
  # remove burnin
  burnin  <-  round(nrow(mc$pars$S)*0.2)
  tmp.pi<-1:nrow(mc$pars$PI)
  id.pi<-tmp.pi[tmp.pi>burnin]
  PI<-mc$pars$PI[id.pi,]
  # estimated PI
  PI_M_SAR<-apply(PI, 2, median)
  PI_M_SAR_M[i,]<-PI_M_SAR
}

# calculate the mean absolute error and worst case error of estimation
PI_ABS_ERROR_SAR  <-  abs(PI_M_SAR_M-PI_TRUE_M)
PI_MAE_SAR  <-  apply(PI_ABS_ERROR_SAR,1,mean)
PI_WCE_SAR  <-  apply(PI_ABS_ERROR_SAR,1,max)

PI_MAE_SAR.df  <-  data.table(PI_MAE=PI_MAE_SAR, REPLICATE=1:100, 
                                       SCENARIO=rep(1,100), N=rep(300,100), SAMP_DIFF=rep(0.15,100))
PI_WCE_SAR.df  <-  data.table(PI_WCE=PI_WCE_SAR, REPLICATE=1:100,
                             SCENARIO=rep(1,100), N=rep(300,100), SAMP_DIFF=rep(0.15,100))

# combine data tables
PI_MAE_SARWS.df[,METHOD:='SARWS']
PI_MAE_SAR.df[,METHOD:='SAR']
PI_MAE.df  <-  rbind(PI_MAE_SARWS.df,PI_MAE_SAR.df)

PI_WCE_SARWS.df[,METHOD:='SARWS']
PI_WCE_SAR.df[,METHOD:='SAR']
PI_WCE.df  <-  rbind(PI_WCE_SARWS.df,PI_WCE_SAR.df)

PI_MAE.df[,ERROR_TYPE:='MAE']
setnames(PI_MAE.df,'PI_MAE','ERROR')
PI_WCE.df[,ERROR_TYPE:='WCE']
setnames(PI_WCE.df,'PI_WCE','ERROR')
PI_ERROR.df <- rbind(PI_MAE.df,PI_WCE.df)
```

The mean absolute error and worst case error of estimated $\pi$ could be loaded through
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
data(twoGroupFlows100_mcmcError)
```

We calculated $95\%$ credibility intervals for mean absolute errors and worst case errors and made a bar plot:
```{r, include=TRUE, eval=FALSE, echo=FALSE, tidy=FALSE, fig.dim = c(8, 3), fig.align="center"}
PI_ERROR.df  <-  twoGroupFlows100_mcmcError[, list(CL = quantile(ERROR,0.025),
                                                   CU = quantile(ERROR,0.975),
                                                   M = quantile(ERROR,0.5)), 
                                            by = c('SAMP_DIFF','SCENARIO','N','METHOD','ERROR_TYPE')]

PI_ERROR.df$ERROR_TYPE  <-  factor(PI_ERROR.df$ERROR_TYPE,levels = c('MAE', 'WCE'),
                                   labels = c('mean absolute error','worst case error'))

library(ggplot2)
ggplot(PI_ERROR.df, aes(x=as.factor(SAMP_DIFF), y=M, fill=METHOD)) +
   geom_bar(stat="identity",position=position_dodge())+
   geom_errorbar(aes(ymin=CL, ymax=CU),width=.2,position=position_dodge(.9))+
   scale_fill_grey()+
	 theme_bw() +
   facet_grid(.~ERROR_TYPE)+
	 labs(x=' sampling differences \n', y='\n error',fill='method')

```
## Data set 2 (more complex, GLM)
We are now ready to consider a more complex example! We are still interested in 
transmission flows between the two groups "1" and "2". However in
real-life sampling may depend on further characteristics of the populations in
the two groups "1" and "2". If this is the case, the above SARWS assumption is
obviously violated. 

**So where do we go from here?** We could consider smaller sub-populations,
though this runs the risk that estimates of the sampling probabilities quickly
become very uncertain. And we know that if the sampling
probabilities are uncertain, the performance of **phyloflows** MCMC engine will
take a big hit. To avoid this problem, let us estimate the sampling
probabilities for each sub-population based on a linear combination of
predictive variables. We call this the GLM approach.


To keep things simple at the start, let us assume that sampling
depends on gender within each group "1" and "2". We consider four 
population groups, "1M", "1F", "2M" and "2F". Let us assume there are $10,000$ individuals
in total, $40\%$ individuals are in location 1, $60\%$ individuals are in
location 2, $60\%$ individuals are women within each location, and that the
sampling rates are $0.6$ for men in group 1, $0.8$ for women in group 1, $0.3$
for men in group 2 and $0.7$ for women in group 2:  

```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE} 
library(data.table)
set.seed(42)
ds <- data.table(SEX=c(rep(c('M','F'),2)),
                 GROUP=c(rep(1,2),rep(2,2)),
                 P_SEQ=c(0.6,0.8,0.3,0.7)) # set up seq rates

ninf <- 1000
ninf1 <- round(ninf*0.4)
ninf2 <- ninf-ninf1
ninf1f <- round(ninf1*0.7)
ninf1m <- ninf1-ninf1f
ninf2f <- round(ninf2*0.6)
ninf2m <- ninf2-ninf2f
ds[,TRIAL:=c(ninf1m,ninf1f,ninf2m,ninf2f)]
ds[,SUC:=rbinom(nrow(ds),TRIAL,P_SEQ)]
ds[,CATEGORY:=paste0(GROUP,':',SEX)]
```

**Next, we calculate the sampling probabilities of transmission flows
within and between the two groups.**  

```{r include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE} 
dobs <- data.table(TR_GROUP=c(rep(1,4),rep(2,4)),
                 REC_GROUP=c(rep(1,2),rep(2,2),rep(1,2),rep(2,2)),
                 TR_SEX=c(rep(c('F','M'),4)),
                 REC_SEX=c(rep(c('M','F'),4)))
dobs[,TR_TRM_CATEGORY:=paste0(TR_GROUP,":",TR_SEX)]
dobs[,REC_TRM_CATEGORY:=paste0(REC_GROUP,":",REC_SEX)]
dobs[, TRM_CAT_PAIR_ID:= seq_len(nrow(dobs))]
set(dobs, NULL, c('TR_GROUP','REC_GROUP','TR_SEX','REC_SEX'), NULL)
tmp <- subset(ds,select=c('CATEGORY','P_SEQ'))
setnames(tmp,colnames(tmp),paste0('TR_TRM_',colnames(tmp)))
dobs <- merge(dobs,tmp,by='TR_TRM_CATEGORY')
setnames(tmp,colnames(tmp),gsub('TR_','REC_',colnames(tmp)))
dobs <- merge(dobs,tmp,by='REC_TRM_CATEGORY')
dobs[, S:= TR_TRM_P_SEQ * REC_TRM_P_SEQ]
set(dobs, NULL, c('TR_TRM_P_SEQ','REC_TRM_P_SEQ'), NULL)
setkey(dobs,TRM_CAT_PAIR_ID)
dobs
```

**Next, we simulate true transmission flows between the sampling groups.**
Let us assume that 30\% and 45\% transmissions occur respectively within the
groups "1" and "2". Further, 10\% are from "1" to "2" and 15\% are from "2" to
"1". Of all those, 60\% originate from men: 
$$
\pi=(0.12,0.18,0.04,0.06,0.06,0.09,0.18,0.27).
$$ 
We further assume the total number of observed transmissions is $N=300$. We
will simulate the actual transmission count $Z$ from a Poisson distribution.
Then we will generate transmission flows between groups by 
$$
z \sim \mbox{Multinomial} (Z,\pi).
$$
Finally we will generate observed transmissions flows by subsampling the
actual transmission flows by
$$
n_{ab} \sim \mbox{Binomial} (z_{ab},\xi_{ab}), \forall a,b,
$$ 
where $\xi_{ab}$ is the probability of sampling a transmission event from $a$ to $b$.

```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE} 
TRUE_PI <- c(0.12,0.18,0.04,0.06,0.06,0.09,0.18,0.27)
N <- rpois(1,300/mean(dobs$S))
z <- rmultinom(1,size=N,prob=TRUE_PI)
n <- matrix(NA_integer_,ncol=1,nrow=length(TRUE_PI))
for (i in 1:length(TRUE_PI)){
	n[i] <- rbinom(1,size=z[i],dobs$S[i])
}
dobs[, TRM_OBS:= n]
dobs<-dobs[TRM_OBS!=0,]
dobs
```

## Corresponding **phyloflows** input data (more complex, GLM)
**We start again by bringing the transmission count data into the form needed
for phyloflows MCMC algorithm.** We add an ID to each observation, called
`TRM_CAT_PAIR_ID`. We also define the sampling groups. As before, we set them to
correspond to the transmission groups even if we are only interested in the
flows within and between groups "1" and "2". After the model is fitted, we
will aggregate the output to the flows between groups "1" and "2". 
```{r, include=TRUE, eval=TRUE,echo=TRUE, tidy=FALSE} 
dobs[, TR_SAMPLING_CATEGORY:=TR_TRM_CATEGORY]
dobs[, REC_SAMPLING_CATEGORY:=REC_TRM_CATEGORY]
setkey(dobs,TR_TRM_CATEGORY,REC_TRM_CATEGORY)
dobs[, TRM_CAT_PAIR_ID:= seq_len(nrow(dobs))]
dobs[, S:=NULL]
dobs
```

**We also need to define the prior distribution on the unknown sampling
probabilities, and generate samples from it**. This final step is where the
GLM approach differs from the SARWS approach. For the 4 groups 
$$
a= 1F, 1M, 2F, 2M
$$
we model the sampling probability by
$$
X_a^i\sim Binom(X_a^i, \xi_a)
$$ 
using the same notation as above, and where 
$$
\mbox{logit}(\xi_{a})=\beta_0+\beta_1 * G_a + \beta_2 S_{a}
$$ 
where $G_a$ equals 0 if the population group is "1", and 1 if the population
group is "2", $S_a$ equals 0 if gender is female and 1 if gender is male, and
the priors on the regression coefficients are quite vague,
$$
\beta_0\sim N(0,100), \beta_1\sim N(0,10), \beta_2\sim N(0,10).
$$
Next we infer the posterior distribution of sampling probabilities with
`stan`. The code for the above GLM regression model is as follows:
```
data{
	int<lower=1> N;
	int SUC[N];
	int TRIAL[N];
	int GROUP[N];
	int MALE[N];
}
parameters{
	real a;
	real grouptwo;
	real male;
}
model{
	vector[N] p_suc;
	male ~ normal( 0 , 10 );
	grouptwo ~ normal( 0 , 10 );
	a ~ normal( 0 , 100 );
	for ( i in 1:N ) {
		p_suc[i] = a + grouptwo * GROUP[i] + male * MALE[i];
	}
	SUC ~ binomial_logit( TRIAL , p_suc );
}
```
Let us fit the model to the sampling data, get samples from the posterior
distributions, and calculate their log-density values: 
```{r include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE,warning=FALSE} 
library(rstan)
nprior <- 1e3
infile.sampling.stan.model <- 'glm_sex2comm2.stan' #in vignettes/glm_sex2comm2.stan
#	run STAN 
tmp <- subset(ds,select=c('SEX','GROUP','TRIAL','SUC'))
tmp[, GROUP:=as.integer(GROUP==2)]
tmp[, MALE:= as.integer(SEX=='M')]
tmp <- as.list(tmp)
tmp$N <- nrow(ds)
fit.par <- stan(  file = infile.sampling.stan.model, 
                  data = tmp, 
                  iter = 2e3,
                  warmup = 5e2,
                  cores = 1,
                  chains = 1,
                  init = list(list(a=0, grouptwo=0, male=0)))

# extract samples from the prior distribution
fit.e <- extract(fit.par)
tmp <- sample(length(fit.e$a), nprior)
ds[, GROUP:=as.integer(GROUP==2)]
ds[, MALE:= as.integer(SEX=='M')]
dprior <- ds[,	
               {
                  z<- with(fit.e, a + grouptwo * GROUP + male * MALE)
                  list(SAMPLE=1:nprior, ETA=as.numeric(z[tmp]))
               },	
               by=c('CATEGORY')]
dprior[, P:= exp(ETA)/(1+exp(ETA))]

# calculate log density using the betakernal bounded density estimator
require(bde)
tmp	<- dprior[, 
		{
			bdest<- bde( P, 
                         dataPointsCache=sort(P), 
                         b=0.001, 
						 estimator='betakernel', 
						 lower.limit=0, 
						 upper.limit=1,
		                 options = list( modified=FALSE, 
								         normalization='densitywise', 
										 mbc='none', 
										 c=0.5))
            list(SAMPLE=SAMPLE, LP=log(density(bdest, P)))
         }, 
		 by='CATEGORY']
dprior <- merge(dprior, tmp, by=c('CATEGORY','SAMPLE'))
set(dprior, NULL, c('ETA'), NULL)	
setnames(dprior, 'CATEGORY', 'SAMPLING_CATEGORY')
head(dprior)
```

This data set can be loaded through
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
data(fourGroupFlows1)
```
