---
title: "04 - Aggregating MCMC output to a new set of variables"
date: "2019-04-30"
output: github_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(knitr)
require(kableExtra)
```


Here we show you how to aggregate estimated transmission flows to a higher
level. Please read the sections "Our Job" and "Our Solution" on the main page
before you go ahead here.

## Getting started

We continue our Very_First_Example from the vignette with the same name here.
The following code chunk contains all code needed, up to running **phyloflows**
MCMC routine. The only change is that the number of iterations is now
$50,000$. The MCMC should take about 2 minutes to run.

```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE,message=FALSE, warning=FALSE} 
require(data.table)
require(phyloflows)
data(twoGroupFlows1, package="phyloflows")
dobs <- twoGroupFlows1$dobs
dprior <- twoGroupFlows1$dprior
mcmc.file <- file.path(getwd(),'twoGroupFlows1_mcmc.RData')
control <- list(seed=42, mcmc.n=5e4, verbose=0)
mc <- phyloflows:::source.attribution.mcmc(dobs, dprior, control)
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, tidy=FALSE, message=FALSE, warning=FALSE}
require(data.table)
require(phyloflows)
data(twoGroupFlows1, package="phyloflows")
dobs <- twoGroupFlows1$dobs
dprior <- twoGroupFlows1$dprior
data(twoGroupFlows1_mcmc, package="phyloflows")
```

## Aggregating flows
Why would it be useful to aggregated the estimated transmission flows? Let us
suppose that group "1" are the individuals aged 15-24 and group "2" are the
individuals aged 25 or older in a population. The estimated flow
vector
$$
\pi=(\pi_{11}, \pi_{12}, \pi_{21}, \pi_{22})
$$
describes the transsmission flow within and between the two age categories. But
what is the overall contribution of transmissions from individuals aged 15-24,
and the overall contribution of transmissions from individuals aged 25+? We
want to estimate
$$ 
\tilde{\pi}= (\tilde{\pi}_{1}, \tilde{\pi}_{2})
$$
where $\tilde{\pi}_{1}=\pi_{11}+\pi_{12}$ and
$\tilde{\pi}_{2}=\pi_{21}+\pi_{22}$. There are many similar scenorios like
that, and **phyloflow** has a little function to help you with that task. The
syntax is as follows.

```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE,message=TRUE, warning=FALSE}
daggregateTo <- subset(dobs, select=c(TRM_CAT_PAIR_ID, TR_TRM_CATEGORY, REC_TRM_CATEGORY))
daggregateTo[, TR_TARGETCAT:= TR_TRM_CATEGORY]
daggregateTo[, REC_TARGETCAT:= 'Any']
set(daggregateTo, NULL, c('TR_TRM_CATEGORY','REC_TRM_CATEGORY'), NULL)	
control <- list( burnin.p=0.05, 
                 thin=NA_integer_, 
				 regex_pars='PI')
mca <- phyloflows:::source.attribution.mcmc.aggregateToTarget(mc=mc, daggregateTo=daggregateTo, control=control)	
mca
```

The output is a data.table that contains the aggregated transmission flows, and
other aggregated variables depending on the value of `control[['regex_pars]]`.
In our case, we removed a burnin-period of 5\% of the MCMC chain, and did not
thin the remaining iterations, yielding about 9500 MCMC samples of the
aggregated flows.

That's it for now. Use your usual R wizadry to process the output further, and
have a look at the other vignettes.

