---
title: "phyloflows: Calculating sources, onward transmissions and flow ratios"
date: "2019-09-10"
author: Xiaoyue Xi and Oliver Ratmann
output: github_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(knitr)
require(kableExtra)
```
```{r, include=FALSE, eval=FALSE, echo=FALSE}
require(rmarkdown)
setwd('/Users/Oliver/git/phyloscanner/phyloflows/vignettes')
infile <- '05_keyquantities.Rmd'
rmarkdown::render(infile, output_format='pdf_document')
rmarkdown::render(infile, output_format='md_document')
```


This vignette describes how a number of key summary statistics of inferred
transmission flows can be easily calculated with **phyloflows**
`source.attribution.mcmc.getKeyQuantities` function. Please work through the
vignette *phyloflows: Estimating transmission flows under heterogeneous sampling
-- a first example* before you go ahead here.

## Getting started

We continue our "First_Example". The following code chunk contains all code
needed, up to running **phyloflows** MCMC routine. The only change is that the
number of iterations is now $50,000$. The MCMC should take about 5 minutes to
run.

```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE,message=FALSE, warning=FALSE} 
require(data.table)
require(phyloflows)
data(twoGroupFlows1, package="phyloflows")
dobs <- twoGroupFlows1$dobs
dprior <- twoGroupFlows1$dprior
mcmc.file <- file.path(getwd(),'twoGroupFlows1_mcmc.RData')
control <- list(seed=42, mcmc.n=5e4, verbose=0)
mc <- phyloflows:::source.attribution.mcmc(dobs, dprior, control)
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, tidy=FALSE, message=FALSE, warning=FALSE}
require(data.table)
# require(phyloflows)
data(twoGroupFlows1, package="phyloflows")
dobs <- twoGroupFlows1$dobs
dprior <- twoGroupFlows1$dprior
data(twoGroupFlows1_mcmc, package="phyloflows")
```

## Sources of transmission for each group
OK, so now we have samples from posterior distribution of transmission flows
within and between the two population groups,
$$
\pi=(\pi_{11},\pi_{12},\pi_{21},\pi_{22}).
$$
One important summary statistic are the sources of transmissions into each
recipient group, defined by
$$
\eta=(\eta_{11},\eta_{21},\eta_{12},\eta_{22})
$$
where $\eta_{ij}= \pi_{ij} / \sum_{s} \pi_{sj}$.

## Onward transmissions from each group
Another important summary statistic are the proportions of transmissions that
originate from each group, defined by 
$$
\nu=(\nu_{11},\nu_{21},\nu_{12},\nu_{22})
$$
where $\nu_{ij}= \pi_{ij} / \sum_{s} \pi_{is}$.

## Transmission flow ratios
Yet another important summary statistic are ratios of transmission flows,
defined by 
$$
\rho_{ij}= \pi_{ij} / \pi_{ji}.
$$

## Calculating key quantities
**phyloflow** has a function to calculate the above
summary statistics. The basic syntax is as follows:
```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE,message=TRUE, warning=FALSE}
#	specify list of user options
#
#	burnin.p: proportion of samples to discard as burn-in 
#			  (only needed when the burn-in was not already removed)
#
#	thin: keep every thin-nth iteration 
#		  (only needed when thinning was not already performed)
#
#	quantiles: quantiles of the marginal posterior distributions 
#			   that will be computed
#
#	flowratios: list of vectors of 3 elements. The 3 elements 
#				specify the name of the flow ratio (first element),
#				the enumerator of the flow ratio (second element),
#				and the denominator of the flow ratio (third element)
control <- list(  burnin.p=0.05, 
			      thin=NA_integer_, 
				  quantiles= c('CL'=0.025,'IL'=0.25,'M'=0.5,'IU'=0.75,'CU'=0.975),
                  flowratios= list( c('1/2', '1 2', '2 1'), c('2/1', '2 1', '1 2'))
                  )
ans <- source.attribution.mcmc.getKeyQuantities(mc=mc, 
		dobs=dobs, control=control)
ans
``` 

## Note
Note it is also possible to specify a file name to MCMC output or
aggregated MCMC output, and it is also possible to input aggregated MCMC output.
Please look up the package help for further instructions.
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE,message=TRUE, warning=FALSE}
?phyloflows::source.attribution.mcmc.getKeyQuantities
``` 