---
title: "06 - Testing phyloflows sampling adjustments"
date: "2019-04-30"
author: Xiaoyue Xi and Oliver Ratmann 
output: github_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include=FALSE} 
library(knitr)
```

```{r, include=FALSE, eval=FALSE, echo=FALSE}
require(rmarkdown)
setwd('~/phyloscanner/phyloflows/vignettes')
infile <- '06_test_sampling_adjustments.Rmd'
rmarkdown::render(infile, output_format='pdf_document')
rmarkdown::render(infile, output_format='md_document')
```

To test the performance of **phyloflows** model for estimating transmission
flows under biased sampling, we set up the following simulation exercise.

## phyloflows MCMC on 100 simulated data sets

First, we simulated several data sets as "Data Set 1 (simple SARWS)". They
can be loaded through: 
```{r include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE}
library(ggplot2)
library(data.table)
library(phyloflows)
data(twoGroupFlows100)
```

We then ran our MCMC algorithm on these 100 data set as follows:
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
for (i in 1:100){
  dobs  <-  twoGroupFlows100[[i]][[1]]
  dprior_sarws  <-  twoGroupFlows100[[i]][[2]]
  dprior_sar  <-  twoGroupFlows100[[i]][[3]]
  randomnumber  <-  twoGroupFlows100[[i]][[4]]
  
  # number of iterations
  tmp  <- subset(dobs, select=c(TRM_CAT_PAIR_ID, TR_SAMPLING_CATEGORY, REC_SAMPLING_CATEGORY))
  tmp  <- melt(tmp, id.vars='TRM_CAT_PAIR_ID', value.name='SAMPLING_CATEGORY', variable.name='WHO')
  dlu            <- unique(subset(tmp, select=c(WHO, SAMPLING_CATEGORY)))
  dlu[, UPDATE_ID:= seq_len(nrow(dlu))]
  
  # mcmc
  mcmc.file  <-  paste0("sarws_mcmc",i,".rda")
  control  <-  list(seed=randomnumber, mcmc.n=nrow(dlu)*1e5, verbose=0, outfile=mcmc.file)
  source.attribution.mcmc(dobs, dprior_sarws, control)
  
  mcmc.file<-paste0("sar_mcmc",i,".rda")
  control<-list(seed=randomnumber, mcmc.n=nrow(dlu)*1e5, verbose=0, outfile=mcmc.file)
  source.attribution.mcmc(dobs, dprior_sar, control)
}
```

## Mean absolute error and worst case error
 
After this, we calculated the mean absolute errors and worst case errors
between the true transmission flows (`TRUE_PI`) and the corresponding mean
estimates of the posterior distribution under **phyloflows** Bayesian
multi-level model with adjustment for sampling heterogeneity.
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
# true PI
TRUE_PI  <-  c(0.36,0.04,0.06,0.54)

# record median of PI for each mcmc outputs
PI_M_SARWS_M <-  matrix(NA_real_,nrow = 100, ncol=4)
PI_M_SAR_M  <-  matrix(NA_real_,nrow = 100, ncol=4)

for (i in 1:100){
  load(paste0("sarws_mcmc",i,".rda"))
  # remove burnin
  burnin  <-  round(nrow(mc$pars$S)*0.2)
  tmp.pi  <-  1:nrow(mc$pars$PI)
  id.pi  <-  tmp.pi[tmp.pi>burnin]
  PI  <-  mc$pars$PI[id.pi,]
  # estimated PI
  PI_M_SARWS  <-  apply(PI, 2, mean)
  PI_M_SARWS_M[i,]  <-  PI_M_SARWS
}

# calculate the mean absolute error and worst case error
PI_TRUE_M  <-  t(replicate(100, TRUE_PI))
PI_ABS_ERROR_SARWS  <-  abs(PI_M_SARWS_M-PI_TRUE_M)
PI_MAE_SARWS  <-  apply(PI_ABS_ERROR_SARWS,1,mean)
PI_WCE_SARWS  <-  apply(PI_ABS_ERROR_SARWS,1,max)
PI_MAE_SARWS.df  <-  data.table(PI_MAE=PI_MAE_SARWS, REPLICATE=1:100,
                                SCENARIO=rep(1,100), N=rep(300,100), SAMP_DIFF=rep(0.15,100))
PI_WCE_SARWS.df  <-  data.table(PI_WCE=PI_WCE_SARWS, REPLICATE=1:100, 
                                SCENARIO=rep(1,100), N=rep(300,100), SAMP_DIFF=rep(0.15,100))

```

Similarly, errors were calculated for the scenario where
no differences appear in sampling. 
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
for (i in 1:100){
  load(paste0("sar_mcmc",i,".rda"))
  # remove burnin
  burnin  <-  round(nrow(mc$pars$S)*0.2)
  tmp.pi<-1:nrow(mc$pars$PI)
  id.pi<-tmp.pi[tmp.pi>burnin]
  PI<-mc$pars$PI[id.pi,]
  # estimated PI
  PI_M_SAR<-apply(PI, 2, median)
  PI_M_SAR_M[i,]<-PI_M_SAR
}

# calculate the mean absolute error and worst case error of estimation
PI_ABS_ERROR_SAR  <-  abs(PI_M_SAR_M-PI_TRUE_M)
PI_MAE_SAR  <-  apply(PI_ABS_ERROR_SAR,1,mean)
PI_WCE_SAR  <-  apply(PI_ABS_ERROR_SAR,1,max)

PI_MAE_SAR.df  <-  data.table(PI_MAE=PI_MAE_SAR, REPLICATE=1:100, 
                                       SCENARIO=rep(1,100), N=rep(300,100), SAMP_DIFF=rep(0.15,100))
PI_WCE_SAR.df  <-  data.table(PI_WCE=PI_WCE_SAR, REPLICATE=1:100,
                             SCENARIO=rep(1,100), N=rep(300,100), SAMP_DIFF=rep(0.15,100))

```


Gathering results together gives data *twoGroupFlows100_mcmcError*.
```{r include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE} 
# combine data tables
PI_MAE_SARWS.df[,METHOD:='SARWS']
PI_MAE_SAR.df[,METHOD:='SAR']
PI_MAE.df  <-  rbind(PI_MAE_SARWS.df,PI_MAE_SAR.df)

PI_WCE_SARWS.df[,METHOD:='SARWS']
PI_WCE_SAR.df[,METHOD:='SAR']
PI_WCE.df  <-  rbind(PI_WCE_SARWS.df,PI_WCE_SAR.df)

PI_MAE.df[,ERROR_TYPE:='MAE']
setnames(PI_MAE.df,'PI_MAE','ERROR')
PI_WCE.df[,ERROR_TYPE:='WCE']
setnames(PI_WCE.df,'PI_WCE','ERROR')
PI_ERROR.df <- rbind(PI_MAE.df,PI_WCE.df)
```


Then we load error data, make plots and compare errors with and without adjustment for sampling heterogeneity. It shows 5\% reduction in mean absolute errors and 10\% reduction in worst case errors on average by adjusting for sampling differences, when it exists (15\%).
```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, fig.dim = c(8, 5), fig.align="center"}
data(twoGroupFlows100_mcmcError)
de <-  twoGroupFlows100_mcmcError[, list( CL = quantile(ERROR,0.025),
                                          CU = quantile(ERROR,0.975),
                                          M = quantile(ERROR,0.5)), 
                                     by = c('SAMP_DIFF','SCENARIO','N','METHOD','ERROR_TYPE')]

de$ERROR_TYPE <- factor(de$ERROR_TYPE,levels = c('MAE', 'WCE'), labels = c('mean absolute error','worst case error'))
ggplot(de, aes(x=as.factor(SAMP_DIFF), y=M, fill=METHOD)) +
   geom_bar(stat="identity",position=position_dodge(.9),width = 0.3)+
   geom_errorbar(aes(ymin=CL, ymax=CU),width=.2,position=position_dodge(.9))+
   scale_fill_grey(start = 0.2, end = 0.8)+
   theme_bw() +
   facet_grid(.~ERROR_TYPE)+
   labs(x=' sampling differences \n', y='\n error',fill='method')
```
