---
title: "07 - Age analysis"
date: "2019-12-10"
author: Xiaoyue Xi and Oliver Ratmann 
output: github_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include=FALSE} 
library(knitr)
require(kableExtra)
```

```{r, include=FALSE, eval=FALSE, echo=FALSE}
require(rmarkdown)
setwd('~/phyloscanner/phyloflows/vignettes')
infile <- '07_age_analysis.Rmd'
rmarkdown::render(infile, output_format='pdf_document')
rmarkdown::render(infile, output_format='md_document')
```

This vignette provides an extension of the general method of **phyloflow**. The aim is to understand the transmission flows between one-year increment age groups. The differences between this example with the general pipeline of **phyloflow** is the correlation between flows. To tackle this problem, we impose a Gaussian process (GP) prior on transmission flows. Also, a more flexible platform, Stan, is introduced to solve the problem. 


# Dataset
**We start with simulating transmission counts** between seven age groups called "15-19","20-24","25-29","30-34","35-39","40-44","45-49". Note that in practice, it would be good to use this method 
to investigate transmission dynamics between one-year increment age group, as the squared exponential kernel is 
for the continuous input space. We first set up hyperparameters for GP, the baseline and sampling fractions.
```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE}
library(rstan)
library(data.table)
library(ggplot2)
library(viridis)
library(phyloflows)
set.seed(42)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
alpha_true <- c(2.5)
rho_true <- c(12,9)
mu_true <- -1
gp_dim <- 2
xi <- c(0.35, 0.45, 0.5, 0.55, 0.5, 0.55, 0.4)
```

**Next, we calculate the sampling probabilities of transmission flows
within and between the two groups.**  

```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE}
dobs <- data.table(expand.grid(TR_TRM_CATEGORY = c("15-19","20-24","25-29","30-34","35-39","40-44","45-49"),
                               REC_TRM_CATEGORY = c("15-19","20-24","25-29","30-34","35-39","40-44","45-49")))
```

```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE}
ds <- data.table(CATEGORY = c("15-19","20-24","25-29",
                              "30-34","35-39","40-44","45-49"),
                 P = xi,ID = 1:7)
```

```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE}
setnames(ds,colnames(ds),paste0('TR_TRM_',colnames(ds)))
dobs <- merge(dobs,ds,by='TR_TRM_CATEGORY')
setnames(ds,colnames(ds),gsub('TR_','REC_',colnames(ds)))
dobs <- merge(dobs,ds,by='REC_TRM_CATEGORY')
setnames(ds,colnames(ds),gsub('REC_TRM_','',colnames(ds)))
dobs[,P:= TR_TRM_P * REC_TRM_P]
dobs[,TR_SMOOTH_CATEGORY:=as.numeric(substr(TR_TRM_CATEGORY,1,2))+2]
dobs[,REC_SMOOTH_CATEGORY:=as.numeric(substr(REC_TRM_CATEGORY,1,2))+2]
dobs[, TR_SAMPLING_CATEGORY:= TR_TRM_CATEGORY]
dobs[, REC_SAMPLING_CATEGORY:= REC_TRM_CATEGORY]
```

**Then we simulate true transmission flows and observed transmission counts.**
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE}
simu_pars <- list(	N=nrow(dobs), D=gp_dim, x=cbind(dobs$TR_SMOOTH_CATEGORY,dobs$REC_SMOOTH_CATEGORY),
                   alpha=alpha_true, rho=rho_true,
                   mu=mu_true,xi=dobs$P)
#	simulate data set	
simu_fit <- stan(	file="simu_poiss.stan", 
                  data=simu_pars, iter=1,
                  chains=1, seed=424838, algorithm="Fixed_param")
dobs$TRM_OBS <- extract(simu_fit)$y[1,]	
```

The data can be loaded through
```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE}
data(sevenGroupFlows1,package='phyloflows')
```

# Input data: observed transmission flows
Input data of the similar format of **phyloflow** are expected.
```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE}
dobs <- subset(dobs, select = c('TR_TRM_CATEGORY', 'REC_TRM_CATEGORY','TR_SAMPLING_CATEGORY',
                                'REC_SAMPLING_CATEGORY', 'TR_SMOOTH_CATEGORY','REC_SMOOTH_CATEGORY',
                                'TRM_OBS'))
head(dobs)
```
**`dobs` specifies observed counts of transmissions from a
transmitter age group to a recipient age group.** It must contain the following columns:
 
* *TR_TRM_CATEGORY* name of transmitter group.
* *REC_TRM_CATEGORY* name of recipient group.
* *TR_SMOOTH_CATEGORY* midpoint of transmitter age group.
* *REC_SMOOTH_CATEGORY* midpoint of recipient age group.
* *TRM_CAT_PAIR_ID* identifier of transmitter-recipient pair
* *TRM_OBS* observed transmission counts

Let us look at the data. The first row shows zero counts of transmission
flows from age group "15-19" to age group "15-19". We visualise our simulated input data in the heatmap : 
```{r data, fig.margin=TRUE}
ggplot(dobs, aes(TR_SMOOTH_CATEGORY, REC_SMOOTH_CATEGORY))+
  geom_tile(aes(fill = TRM_OBS)) +
  scale_fill_viridis() +
  scale_x_continuous(expand = c(0,0), limits = c(14.5,49.5))+
  scale_y_continuous(expand = c(0,0), limits = c(14.5,49.5))+
  theme_classic()+
  labs(x='transmitter age group \n', y='\n recipient age group',fill='transmission \n counts')

```

## Input data: sampling information
**`dobs` also requires information about how each group was sampled. **
This is stored in the following columns:

* *TR_SAMPLING_CATEGORY* sampling strata of transmitter group
* *REC_SAMPLING_CATEGORY* sampling strata of recipient group

**`dprior.fit` specifies the distribution of sampling probability in each sampling group.** 
The sampling probability is either characterised by beta-binomial model or GLM model (See 01 - Simulating data). and given in the form of samples. In the tutorial, we opted for a probabilistic programming language Stan, and it cannot take samples as input. However, it is possible to fit statistical distributions to samples and input distribution parameters. The information is stored in data *dprior.fit* in the following columns: 

* *SAMPLING_CATEGORY* name of sampling strata
* *ALPHA, BETA* shape parameters of the distribution of sampling probability.

Under beta-binomial model, the posterior distribution of the sampling probability is known analytically, i.e. beta. Let us look at the sampling information:


```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=FALSE, warning=FALSE}
ds$TRIAL <- c(4000, 3700, 3300, 2500, 1700, 1000, 500)
ds[,SUC := round(TRIAL * P)]
dprior.fit <- copy(ds)
dprior.fit[,ALPHA := SUC+1]
dprior.fit[,BETA := TRIAL-SUC+1]
dprior.fit <- subset(dprior.fit, select = c('CATEGORY','ALPHA','BETA'))
colnames(dprior.fit) <- c('SAMPLING_CATEGORY','ALPHA','BETA')
head(dprior.fit)
```



# Method
Again we use a Bayesian approach to estimate the proportion of
transmissions between the two population groups. Please see the problem setting in 
**phyloflows: Estimating transmission flows under heterogeneous sampling -- a first example**.
Unlike the simple example,
a probabilistic programming language Stan provides a flexible platform to impose
different kinds of priors.
Recall the posterior distribution of the parameters $(\lambda, s)$ is given by 
$$
\begin{aligned}
p(\lambda, s | n) & \propto p(n | \lambda, s) p(\lambda, s) \\
              & = \prod_{i=1,\cdots,7;j=1\cdots,7} Poisson(n_{ij};\lambda_{ij}*s_i*s_j) p(\lambda_{ij}) p(s_i) p(s_j).
\end{aligned}
$$
Then, we calculate the main quantity of interest, $\pi$, via
$$
\pi_{ij}= \lambda_{ij} / \sum_{k=1,\cdots,7;  l=1,\cdots,7} \lambda_{kl}.
$$ 
for $i=1,\cdots,7$ and
$j=1,\cdots,7$.

**For the prior distributions**, we specify for $p(\lambda_{ij})$, $i=1,2; j=1,2$ 
uninformative prior distributions. We use a Gamma distribution with parameters
$\alpha_i=0.8/7^2$ and $\beta=0.8/Z$ with
  $Z= \sum_{ij | n_{ij}>0} n_{ij}/(s_i*s_j) + \sum_{ij | n_{ij}>0}
(1-s_i*s_j)/(s_i*s_j)$. This choice implies for $\pi$ a Dirichlet prior distribution with parameters $\alpha_i$, which
  is considered to be an objective choice. For $p(s_i)$, we use a strongly informative
prior distribution, based on the available data as illustrated above.

An alternative prior distribution is Gaussian process prior, which penalises large changes between neighbouring age groups, and the kernel $k$ is given by 

$$
k((a,b),(a',b'))= \sigma^2 \exp(-0.5 ((\frac{a-a'}{\ell_1})^2 + (\frac{b-b'}{\ell_2})^2 ))
$$
That is, $\lambda_{i,j} = \mu + f$ where $f\sim \mathcal{GP}(0,k)$.

# MCMC
## MCMC syntax 
We use a Markov Chain Monte Carlo algorithm to sample from the posterior
The syntax for running the algorithm is as follows. 
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE}
# specify a list of control variables:
#	seed  	random number seed
#	mcmc.n 	number of MCMC iterations
#	method "gamma" or "gp"
#	outfile output file name if you like to have the results 
#			written to an *.rda* file
control <- list(seed=42, mcmc.n=500, method='gamma')
# run MCMC
ans.gamma <- source.attribution.mcmc.stan(dobs, dprior.fit, control)
# save(ans.gamma, file = '~/phyloscanner/phyloflows/data/sevenGroupFlows1_gamma_mcmc.RData')

control <- list(seed=42, mcmc.n=500, method='gp')
# run MCMC
ans.gp <- source.attribution.mcmc.stan(dobs, dprior.fit, control)
# save(ans.gp, file = '~/phyloscanner/phyloflows/data/sevenGroupFlows1_gp_mcmc.RData')
```

```{r, include=FALSE, eval=TRUE, echo=TRUE, tidy=FALSE, message=TRUE}
data(sevenGroupFlows1_gamma_mcmc)
data(sevenGroupFlows1_gp_mcmc)
```

## MCMC output
Let us have a look at the output:  
  ```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE, message=TRUE}
str(ans.gamma)
str(ans.gp)
```

We are mostly interested in the marginal posterior distribution 
$$
  p(\pi | n ),
$$
and the algorithm returns samples of the log Poisson rates $\log \lambda$, 
from which the transmission flows $\pi$ can be calculated.



In addition, estimated parameters can be compared with the true hyperparameter values.
```{r posterior, fig.margin=TRUE}
params_gamma <- extract(ans.gamma)
params_gp <- extract(ans.gp)
c_light <- c("#DCBCBC")
c_dark <- c("#8F2727")
c_dark_highlight <- c("#7C0000")
par(mfrow=c(2, 2))	
hist(params_gp$alpha, main="", xlab="alpha", col=c_dark, border=c_dark_highlight, yaxt='n')
abline(v=2.5, col=c_light, lty=1, lwd=3)
hist(params_gp$rho[,1], main="", xlab="rho1", col=c_dark, border=c_dark_highlight, yaxt='n')
abline(v=12, col=c_light, lty=1, lwd=3)
hist(params_gp$rho[,2], main="", xlab="rho2", col=c_dark, border=c_dark_highlight, yaxt='n')
abline(v=9, col=c_light, lty=1, lwd=3)
hist(params_gp$mu, main="", xlab="mu", col=c_dark, border=c_dark_highlight, yaxt='n')
abline(v=-1, col=c_light, lty=1, lwd=3)
```


That's it for now. Of course we would like to run the MCMC chain for longer,
perhaps 10,000 sweeps. To check for convergence, calculate
effective sample sizes, and quantiles of the posterior distribution, 
have a look at the other vignettes.






```{r, include=FALSE, eval=FALSE, echo=FALSE, tidy=FALSE, message=FALSE, warning=FALSE}
range(summary(fit.gamma)$summary[, "n_eff"])
range(summary(fit.gamma)$summary[, "Rhat"])
range(summary(fit.gp)$summary[, "n_eff"])
range(summary(fit.gp)$summary[, "Rhat"])
```